`phyr`: An R package for phylogenetic species-distribution modelling in ecological communities

### 2019-11-14 ms.Rmd
update ms [skip ci]

```{r knitr_options, include = FALSE}
if(!require("xfun")) install.packages("xfun")
if(!require("phyr")) devtools::install_github("daijiang/phyr")
xfun::pkg_attach2(c("phyr", "knitr", "kableExtra", "kableExtra", "tidyverse"))
# opts_knit$set(root.dir = normalizePath("../"))
opts_chunk$set(fig.align='center', cache = FALSE, warning = FALSE, 
               message = FALSE, echo = TRUE)
options(digits = 3, width = 88, knitr.kable.NA = '')
# traits, phy, dat
source(file = "simulation_pglmm.R", echo = FALSE, print.eval = FALSE)
```

@@

1. Model-based approaches are getting more and more popular in ecological studies, with joint species distribution models as a particular example. However, most of such studies did not include phylogenetic relationships among species despite its well-known importance in shaping species distribution and community composition. One potential reason is the lack of tools to allow ecologists to fit phylogenetic species distribution models easily.
2. To fill this gap, the R package `phyr` provides a general formula syntax and implementations of phylogenetic generalized linear mixed models. It also provides updated functions to calculate community phylogenetic diversity and correlations among functional traits while accounting for phylogenetic relationships.
3. Other programs can also be used to fit phylogenetic species distribution models, though they require considerable programming and expertise. The model formula syntax we proposed here can be applied to those programs and thus serves as a unified framework for phylogenetic species distribution modelling.
4. `phyr` can help current and future researchers formulate and analyze phylogenetic species-distribution models.

_Keywords_: phylogenetic diversity, phylogenetic generalized linear mixed models, functional trait, trait correlation, model-based methods

@@

Ecological communities are collections of species that occur within the same geographical area. Which species occur within communities depends on the dispersal ability of species to enter the community, the environmental conditions that they find there, and the interactions that they have with other species in the community. These three processes -- dispersal, environmental tolerance, and species interactions -- depend on the traits that species possess and hence reflect evolutionary history and biogeographic processes [@warren2014mistaking; @gerhold2018deep]. For example, an insect species might only occur in a lake if its adult stage has long-distance flight capabilities, if it can tolerate the low pH of the lake, and if it can avoid the predators that are common. Because traits play a central role in the composition of species that make up a community, community composition will likely reflect, at least in part, phylogenetic relationships among species. For example, two closely related insects might have similar dispersal capability, pH tolerance, and predator avoidance behavior, making them more likely to occur in the same lake. The recognition that phylogenetic relationships can increase our understanding of communities has led to a growing number of statistical methods for analyzing phylogenetic community composition [@ives2011generalized; @li2017can; @frishkoff2017phylogenetic].

Just as the distributions of two species might reflect their proximity on a phylogenetic tree, the species occurring at two sites might reflect the sites’ geographical proximity. The most immediate possible cause of spatial correlations in species distributions is dispersal, if nearby sites are more likely to be colonized by a species. Spatial proximity may also be a surrogate for environmental variables that are unknown or unmeasured. For example, an insect species might occur in two nearby lakes because they both have low pH, yet pH has not been measured. Just as phylogenetic relationships among species can generate correlations between species in which sites they occupy, so too can spatial proximity generate correlations between sites in the species they contain [@ives2006statistics].

How species respond to environmental factors, and how they respond to each other, depend on their traits. Therefore, relationships among functional traits can provide insights about the evolutionary history that has shaped species trait so they can occupy the same sites. For example, two insect species that occur in the same lake might share both long-range flight abilities and tolerance to low pH. Is the positive correlation between these two traits caused by correlated selective forces? A challenge to answering this question is that phylogenetic correlations between trait values of closely related species can give the appearance of positive correlations between traits: two species might have both long-range flight abilities and tolerance to low pH only because they are phylogenetically closely related. Therefore, it is important to account for species’ evolutionary history when studying correlations among functional traits.

Statistical models for phylogenetic community composition provide flexible tools for exploring the many possible factors underlying the distribution of species and the composition of communities [@ives2011generalized]. The models can describe complex relationships in the data, such as how phylogenetically related species might respond similarly to the same environmental gradient, or how phylogenetically related species might exclude each other from the same communities. They also give a firm statistical basis to test these patterns, the ability to simulate data sets from the fitted model, and the ability to predict the composition of unsurveyed communities. These benefits of phylogenetic community composition models come with costs: building models can be intricate and fitting them computationally slow. The R package `phyr` is designed to overcome many of the costs with a user-friendly interface, flexibility to build a rich collection of models, and good computational performance. Below, we first give a brief overview of the structure and syntax of two key functions `pglmm()` and `cor_phylo()`. We then compare them to methods and programs that are currently available. Finally, we applied `pglmm()` and `cor_phylo()` to simulated data to illustrate their implementation and output.

@@

`Phyr` contains three groups of functions (Table \@ref(tab:mainFunc)): phylogenetic generalized linear mixed models (`pglmm()`), phylogenetic correlations between functional traits (`cor_phylo()`), and community phylogenetic diversity metrics (e.g. `psv()`, `pse()`). The workhorse functions of all groups are written in C++ to increase computational speed. Here, we will focus on the first two groups of functions because they are more complicated than community phylogenetic diversity metrics.

@@

`pglmm()` constructs and fits generalized linear mixed models that incorporate covariance matrices containing the phylogenetic relationships among species. The syntax for `pglmm()` resembles that used in the R package `lme4` [@bates2015lme4], and indeed `pglmm()` will fit most of the models that can be fit with `lmer()` and `glmer()`. `pglmm()` goes beyond `lmer()`and `glmer()` by allowing the specification of covariance matrices, which could be a phylogenetic covariance matrix or any other covariance matrix that the user defines (e.g., spatial or temporal autocorrelation matrix). `pglmm()` can also fit models with “nested” covariance structures (e.g. species phylogenetic covariance matrix nested within site covariance matrix). `pglmm()` can operate in both frequentist mode, with the distribution of species among communities being Gaussian, Binary, Binomial or Poisson, and Bayesian mode with the addition of zero-inflated Binomial and Poisson distributions. Finally, it is our hope that the formula syntax of `pglmm()` can be used to fit similar models with other programs such as Stan (e.g. via R package `brm` [@brmsR2018]).

@@

Here, Y is a binary (Bernoulli) dependent variable which takes values of either 0 or 1. The specification `family = 'binomial'` allows binary data and also binomial data for which Y is a matrix containing columns for successes and failures. The independent variables `trait` and `env` take on different values for each species and site, respectively. Sites and species (`sp`) are treated as random effects: thus, `(1|site)` implies that a value from a Gaussian random variable is picked for each site, thereby representing unmeasured differences among sites. For the case of species, the double underscore in `(1|sp__)` implies that, in addition to a random effect for species, there is a second random effect which contains the phylogenetic relationships among species. The phylogenetic random effect assumes that values for each species are picked from a multivariate Gaussian distribution with phylogenetic covariance matrix 𝚺. A covariance matrix 𝚺 is specified by `cov_ranef = list(sp = phy.sp, site = V.space)`. The covariance matrix `phy.sp` associated with species can be a `phylo` object from the R package `ape` [@ape2018]. It is also possible to specify an explicit covariance matrix, such as `site = V.space`. To construct 𝚺 from a 'phylo' object, `pglmm()` assumes that the residual variation associated with species follows a Brownian motion model of evolution, so that the covariance between two species is proportional to the distance between them on the phylogenetic tree.

The syntax `(1|sp__)` or  `(1|site__)` generates two random effects, one without and one with phylogenetic or spatial covariances; in contrast, `(1|sp)` would generate only a single random effect that is independent among species. `pglmm()` forces in a term for `(1|sp)` whenever `(1|sp__)` is specified, because otherwise any difference among species would be captured by the diagonal elements in 𝚺 even in the absence of covariances among phylogenetically related species which are specified by the off-diagonal elements of 𝚺. Therefore, if `(1|sp)` were not included, this could lead to the identification of phylogenetic signal in the abundances of species even in its absence. To account for differences among sites in how they select for species with different traits, `(trait|site)` allows the slope of Y against `trait` to be a Gaussian random variable. Similarly, to account for the differences among species for how they respond to `env`, `(env|sp__)` allows the relationship of Y against `env` to be given by two slopes, the first slope that is picked from a Gaussian random variable in which species are independent and the second slope that is picked from a multivariate Gaussian with covariance matrix 𝚺. Finally, `(1|sp__@site)` generates a nested term: within a site, the residual variation in Y shows phylogenetic relatedness, with phylogenetically related species more likely to occur in the same site. Note that `(1|sp__)` differs from `(1|sp__@site)` because `(1|sp__)` generates differences in the mean value of Y for species across all sites, whereas `(1|sp__@site)` is local, giving the covariances among species only within sites. This nested term can be used to test for community clustering or overdispersion [@webb2002phylogenies; @ives2011generalized]. Other forms of a nested term are available in `pglmm()`, which can be used to study more complicated questions such as bipartite networks.

@@

```{r designPlot, fig.cap="(ref:designPlotCap)", out.width="100%"}
knitr::include_graphics(normalize_path("designPlot.png"))
```

@@

`cor_phylo()` makes it possible to compare suites of traits among species, accounting for their phylogenetic relatedness. To identify suites of traits under joint selection, such as traits that together make up adaptive syndromes, it is necessary to perform a correlation analysis in which phylogenetic relatedness is factored out. `cor_phylo()` does this. It can also include within-species variation (e.g., measurement error) which should better-expose the underlying correlations in traits among species. Whereas `pglmm()` can be used to identify the composition of communities within a region, `cor_phylo()` can be used to assess patterns of traits among species that make up the regional species pool.

@@

In this example, the correlation between two traits, `trait1` and `trait2`, is assessed, and the column named `sp` in `data` identifies the species. The object `phy.sp` specifies the phylogenetic covariance matrix as a 'phylo' object from the `ape` package. `cor_phylo()` estimates the phylogenetic signal for each trait by assuming that trait evolution is given by a Ornstein-Uhlenbeck process. The term `covariates = list(trait1 ~ env)` includes the independent variable `env` for `trait1`, to remove possible confounding effects; only an intercept is estimated if no covariate is provided for a trait. Within-species variation (measurement error) is specified by `meas_errors = list(trait1 ~ me1, trait2 ~ me2)`, where `me1` and `me2` are the standard errors for `trait1` and  `trait2`, respectively, of values at the tips of the phylogenetic tree. If within-species standard errors are not provided for a given trait, the trait values are assumed to be known without error. Finally, `cor_phylo()` can perform parametric bootstrapping to give confidence intervals for all parameter estimates: correlations, phylogenetic signals, covariate coefficients, and coefficient covariances.

@@

`pglmm()` is syntactically modeled after `lmer()` and `glmer()` in `lme4`, although it allows the specification of phylogenetic covariance matrices. `pglmm()` also allows "nested" models which arise when phylogenetic covariances only act within single communities, rather than among communities. Such nested models make it possible to assess whether phylogenetic relatedness affects the abundance of species within the same communities, such as whether competition between closely related species excludes one of the competitors from communities where the other is present. Nested models are structurally incompatible with the architecture of `lme4`.

There are alternative programs to `pglmm()`, although they have limitations that `pglmm()` overcomes. @hadfield2013tale use the R package `MCMCglmm` [@mcmcglmm2010] to perform phylogenetic community analyses, although they also use ASReml because its penalized quasi-likelihood (PQL) approach is computationally much faster. Hierarchical Modelling of Species Communities (HMSC-R) [@tikhonov2019joint] performs community analyses using Bayesian MCMC approaches, although it does not include nested terms. It is also possible to code up specific phylogenetic community models using flexible Bayesian platforms such as WinBugs, Stan, and JAGS, although this will involve considerable programming and expertise.

@@

Joint Species Distribution Models (JSDMs) are models where the response is distribution (abundances or occurrences) of multiple species across sites or samples, where all species are modelled jointly, usually by allowing non-zero covariance between either species-level errors, species-level coefficients in the model, or both [@warton2015so]. A recent review and evaluation of JSDMs, for example, only considers models that allow modelling residual dependencies among species [@wilkinson2019comparison].

`pglmm()` in `phyr` is a joint species distribution model where the (residual) dependencies among species are modelled as a simple function of phylogeny. JSDMs, and Species Distribution Models (SDM) in general, have typically been focused on producing accurate predictions of how species are distributed, usually in a geographic context. However, they can also be used for making inferences about the biology of species, such as which environmental factors are important in a species or set of species' biology, or whether traits or evolutionary history can help explain the structure of species communities. It is this kind of inference that is the focus of the JSDM implemented in `pglmm()`. There is often a trade-off between improving predictions and making solid inferences, because increasing the complexity or flexibility of a model can improve its predictive power, but this same complexity makes it more difficult to understand what biology is being represented by the model outputs. In `pglmm()`, the focus is on understanding whether evolutionary history is important in structuring species assemblages, which is facilitated by using a phylogenetic covariance model. This is in contrast to many other JSDMs [e.g. those in described in @wilkinson2019comparison], which attempt to flexibly estimate all pairwise species covariances, without taking into account any relationships amongst species, in an attempt to maximize predictive power.

Of course, the goals of solid inference and prediction are not mutually exclusive. Good prediction requires avoiding overfitting, which can be facilitated by reducing the number of parameters in the model. In some systems, it may be possible to make better predictions using a simple phylogenetic model if phylogeny is a strong predictor of species covariance, or if many species are poorly sampled and thus estimating covariances between them individually results in higher prediction variance. Using phylogeny can help closely related species share statistical strength through phylogenetic partial pooling. Ultimately, it can be very powerful to use the same kind of statistical framework to do both predictive and inferential work in ecology, because it allows ecologists to smoothly move between these two goals, both of which are highly important in modern ecology, and more easily and quickly draw mutual insights between them.

@@

We simulated datasets to demonstrate how to use `pglmm()` and `cor_phylo()`. Details about simulations can be found in the Appendix. Our goal of this section is to provide some general ideas about the input and output of these two functions instead of testing their statistical performances or interpreting the ecological meanings of model results. For those purposes, please see the package vignettes and @ives2018book.

@@

We fitted a PGLMM that examined how a hypothetical functional trait, environmental gradient, and their interaction affect distributions of `r nspp` species across `r nsite` sites (we focused on abundance and used the default family of data distribution [Gaussian] but other distributions can also be specified by setting the `family` argument) while accounting for species’ phylogenetic relationships and site spatial autocorrelations (by setting `cov_ranef = list(species = phy, site = Vspace)` where `species` and `site` are group variables of random terms, `phy` can be a phylogeny with class `phylo` or a phylogenetic variance-covariance matrix, `Vspace` is a matrix of spatial distances among sites). This model can also be fitted with Bayesian framework by setting `bayes = TRUE`.

@@

The summary of model results includes the model fitting method (maximum likelihood or bayesian), the model formula, log likelihood and other statistics based on it (e.g., AIC, BIC, DIC), estimates of variances of random terms, coefficients of fixed terms and their uncertainties (Standard errors, Z scores, and P values if fitted by maximum likelihood or lower and upper boundaries of credible intervals).

@@

Here, we simulated two hypothetical functional traits (`trait_1` and `trait_2`) of 50 species. We set the true correlation between these two traits to be 0.7 and their phylogenetic signals to be 0.3 and 0.95, respectively.

@@

The output of `cor_phylo()` includes log likelihood values, AIC, BIC, estimated correlation matrix of traits, estimated phylogenetic signals of traits, estimated coefficients and their uncertainties (standard errors, Z scores, and p values). If bootstrapping was enabled by setting the `boot` argument, the lower and upper boundaries of correlations, phylogenetic signal values, and coefficients will be appended.

@@

In recent years, there has been increasing effort to apply model-based approaches in community ecology. One particular example is the joint species distribution models. Despite the well-known importance of phylogenetic relationship in structuring species distributions and community composition, relatively few studies have incorporated phylogenetic relationships in model-based analyses of species distributions and community ecology. One potential reason is the lack of easy-to-use tools to facilitate the usage of phylogenetic species-distribution modelling in ecological communities. The package `phyr` fills this gap by providing implementations of phylogenetic species-distribution models with flexible model formula syntax (`pglmm()`). It also includes other model-based functions that are useful for ecological studies such as calculating community phylogenetic diversity (e.g. `psv()`) and estimating correlations among functional traits while accounting for their evolutionary history (`cor_phylo()`) (Table \@ref(tab:mainFunc)).

The model formula of `pglmm()` is general and can be applied using other tools to fit phylogenetic species-distribution models. Thus, `pglmm()` can serve the developer community as a shell for new methods that fit GLMMs, with `phyr` providing an easy user interface. We hope that the `phyr` package will help current and future researchers formulate and analyze phylogenetic species-distribution models.

@@

No data were used in this study. `phyr` is available at Github (https://github.com/daijiang/phyr) and will be submitted to CRAN in the near future.

>>>>

```{r knitr_options, include = FALSE}
if(!require("xfun")) install.packages("xfun")
if(!require("phyr")) devtools::install_github("daijiang/phyr")
xfun::pkg_attach2(c("phyr", "knitr", "kableExtra", "kableExtra", "tidyverse"))
# opts_knit$set(root.dir = normalizePath("../"))
opts_chunk$set(fig.align='center', cache = TRUE, warning = FALSE, 
               message = FALSE, echo = TRUE)
options(digits = 3, width = 88, knitr.kable.NA = '')
# traits, phy, dat
source(file = "simulation_pglmm.R", echo = FALSE, print.eval = FALSE)
```

@@

1. Model-based approaches are increasingly popular in ecological studies. A good example of this trend is the use of joint species distribution models to ask questions about ecological communities. However, most current applications of model-based methods do not include phylogenies despite the well-known importance of phylogenetic relationships in shaping species distributions and community composition. In part, this is due to lack of accessible tools allowing ecologists to fit phylogenetic species distribution models easily.
2. To fill this gap, the R package `phyr` provides implementations of phylogenetic generalized linear mixed models. It also provides functions to calculate community phylogenetic diversity, correlations among functional traits while accounting for phylogenetic relationships, and generalized models for analyzing phylogenetic comparative data.
3. Other programs can also be used to fit phylogenetic species distribution models, but they require considerable programming and expertise. The model formula syntax we propose serves as a simple and unified framework for phylogenetic species distribution modelling.
4. `phyr` can help current and future researchers formulate and analyze phylogenetic joint species-distribution models.

_Keywords_: phylogenetic diversity, phylogenetic generalized linear mixed models, functional trait, trait correlation, model-based methods, Joint Species Distribution Model

@@

Ecological communities are collections of species that occur within the same geographical area. Which species occur within communities depends on the dispersal ability of species to enter the community, the environmental conditions that they find there, and the interactions that they have with other species in the community. These three processes -- dispersal, environmental tolerance, and species interactions -- depend on the traits that species possess and hence reflect evolutionary history and biogeographic processes [@warren2014mistaking; @gerhold2018deep]. For example, the larvae of an aquatic insect species might only occur in a lake if its adult stage has long-distance flight capabilities, if it can tolerate the low pH of the lake, and if it can avoid the predators that are common. Because traits play a central role in the composition of species that make up a community, community composition will likely reflect, at least in part, phylogenetic relationships among species. For example, two closely related insects might have similar dispersal capability, pH tolerance, and predator avoidance behavior, making them more likely to occur in the same lake. The recognition that phylogenetic relationships can increase our understanding of communities has led to a growing number of statistical methods for analyzing phylogenetic community composition [@ives2011generalized; @li2017can; @frishkoff2017phylogenetic; @losos1996; @webb2000; @webb2002phylogenies; @cavender2004; @helmus2007].

Just as the distributions of two species might reflect their proximity on a phylogenetic tree, the species occurring at two sites might reflect the sites’ geographical proximity. The most immediate possible cause of spatial correlations in species distributions is dispersal, if nearby sites are more likely to be colonized by a species. Spatial proximity may also be a surrogate for environmental variables that are unknown or unmeasured. For example, an insect species might occur in two nearby lakes because they both have low pH, yet pH has not been measured. Just as phylogenetic relationships among species can generate correlations between species in which sites they occupy, so too can spatial proximity generate correlations between sites in the species they contain [@cressie1991; @ives2006statistics].

How species respond to environmental factors, and how they respond to each other, depend on their traits. Therefore, correlations among functional traits can provide insights about the evolutionary history that has shaped species traits so they can occupy the same sites. For example, two insect species that occur in the same lake might share both long-range flight abilities and tolerance to low pH. Is the positive correlation between these two traits caused by correlated selective forces? A challenge to answering this question is that phylogenetic correlations between trait values might reflect species phylogenetic relatedness rather than shared selection: two species might have both long-range flight abilities and tolerance to low pH only because they are phylogenetically closely related. To distinguish between these two explanations -- convergence of suites of traits due to shared selective forces versus similarity due to phylogenetic relatedness -- it is necessary to account for phylogenies when performing correlation analyses between traits that could explain similarities in the distributions of species.

Statistical models for phylogenetic community composition provide flexible tools for exploring the many possible factors underlying the distribution of species and the composition of communities [@ives2011generalized; @ovaskainen2011; @warton2015so]. The models can describe complex relationships in the data, such as how phylogenetically related species might respond similarly to the same environmental gradient, or how phylogenetically related species might exclude each other from the same communities. They also give a firm statistical basis to test these patterns, the ability to simulate data sets from the fitted model, and the ability to predict the composition of unsurveyed communities. These benefits of phylogenetic community composition models come with costs: building models can be intricate and fitting them computationally slow. The R package `phyr` is designed to overcome many of these costs with a user-friendly interface, flexibility to build a rich collection of models, and good computational performance. Below, we first give a brief overview of the structure and syntax of two key functions `pglmm()` and `cor_phylo()`. `pglmm()` allows the formulation of a diverses set of phylogenetic generalized linear models that can be used not only to analyze phylogenetic community composition but also comparative models for Gaussian and non-Gaussian data. `cor_phylo()` computes the Pearson correlations among species traits while simultaneously estimating the strength of phylogenetic signal within each trait. We then compare `pglmm()` and `cor_phylo()` to methods and programs that are currently available. Finally, we apply  `pglmm()` and `cor_phylo()` to simulated data to illustrate their implementation and output.

@@

`Phyr` contains three groups of functions (Table \@ref(tab:mainFunc)): phylogenetic generalized linear mixed models (`pglmm()`), phylogenetic correlations between functional traits (`cor_phylo()`), and community phylogenetic diversity metrics (e.g., `psv()`, `pse()`). The workhorse functions of all groups are written in C++ to increase computational speed. Here, we will focus on the first two groups of functions, because they are more complicated and less readily available to practitioners than community phylogenetic diversity metrics.

@@

`pglmm()` constructs and fits generalized linear mixed models that incorporate covariance matrices containing the phylogenetic relationships among species. The syntax for `pglmm()` resembles that used in the R package `lme4` [@bates2015lme4], and indeed `pglmm()` will fit most of the models that can be fit with `lmer()` and `glmer()`. `pglmm()` goes beyond `lmer()`and `glmer()` by allowing the specification of covariance matrices, which could be phylogenetic covariance matrices or any other covariance matrices that the user defines (e.g., spatial or temporal autocorrelation matrix). `pglmm()` can also fit models with “nested” covariance structures (e.g., a species phylogenetic covariance matrix nested within a site covariance matrix). `pglmm()` can operate in both frequentist mode, with the distribution of species among communities being Gaussian, binary, binomial or Poisson, and Bayesian mode with the addition of zero-inflated binomial and Poisson distributions. Finally, it is our hope that the formula syntax of `pglmm()` can be used to fit similar models with other programs such as Stan (e.g. via R package `brms` [@brmsR2018]).

@@

Here, Y is a binary (Bernoulli) dependent variable which takes values of either 0 or 1. The specification `family = 'binomial'` allows binary data and also binomial data for which Y is a matrix containing columns for successes and failures. The independent variables `trait` and `env` take on different values for each species and site, respectively. Sites and species (`sp`) are treated as random effects: `(1|site)` implies that a value from a Gaussian random variable is picked for each site, thereby representing unmeasured differences among sites. For the case of species, the double underscore in `(1|sp__)` implies that, in addition to a random effect for species, there is a second random effect which contains the phylogenetic relationships among species. The phylogenetic random effect assumes that values for each species are picked from a multivariate Gaussian distribution with phylogenetic covariance matrix 𝚺. A covariance matrix 𝚺 is specified by `cov_ranef = list(sp = phy.sp, site = V.space)`. The covariance matrix `phy.sp` associated with species can be a `phylo` object from the R package `ape` [@ape2018]. It is also possible to specify an explicit covariance matrix, such as `site = V.space`. To construct 𝚺 from a 'phylo' object, `pglmm()` assumes that the residual variation associated with species follows a Brownian motion model of evolution, so that the covariance between species is proportional to their shared evolutionary history (e.g., shared branch length on a phylogeny).

The syntax `(1|sp__)` or  `(1|site__)` generates two random effects, one without and one with phylogenetic or spatial covariances; in contrast, `(1|sp)` would generate only a single random effect that is independent among species. `pglmm()` forces in a term for `(1|sp)` whenever `(1|sp__)` is specified, because otherwise any difference among species would be captured by the diagonal elements in 𝚺 even in the absence of covariances among phylogenetically related species which are specified by the off-diagonal elements of 𝚺. Therefore, if `(1|sp)` were not included, this could lead to the identification of phylogenetic signal in the abundances of species even in its absence. To account for differences among sites in how they select for species with different traits, `(trait|site)` allows the slope of Y against `trait` to be a Gaussian random variable. Similarly, to account for the differences among species for how they respond to `env`, `(env|sp__)` allows the relationship of Y against `env` to be given by two slopes, the first slope that is picked from a Gaussian random variable in which species are independent and the second slope that is picked from a multivariate Gaussian with covariance matrix 𝚺. Finally, `(1|sp__@site)` generates a nested term: within a site, the residual variation in Y shows phylogenetic relatedness, with phylogenetically related species more likely to occur in the same site. Note that `(1|sp__)` differs from `(1|sp__@site)` because `(1|sp__)` generates differences in the mean value of Y for species across all sites, whereas `(1|sp__@site)` is local to sites, giving the covariances among species only within sites. This nested term can be used to test for community clustering or overdispersion [@webb2002phylogenies; @ives2011generalized]. Other forms of a nested term are available in `pglmm()`, which can be used to study more complicated questions such as bipartite networks.

@@

```{r designPlot, fig.cap="(ref:designPlotCap)", out.width="100%"}
knitr::include_graphics(normalize_path("designPlot2.png"))
```

Whereas `pglmm()` is designed to accept community composition data, in which the same species can occur in multiple sites, the algorithm used by `pglmm()` can equally be used for comparative data in which each species is represented by only a single data point. `pglmm.compare()` is a wrapper for `pglmm()` that is tailored for comparative data and thus provides an easy-to-use function for analyzing non-Gaussian phylogenetic data.

@@

`cor_phylo()` makes it possible to compare suites of traits among species, accounting for their phylogenetic relatedness [@zheng2009new; @johnson2014]. To identify suites of traits under joint selection, such as traits that together make up adaptive syndromes, it is necessary to perform a correlation analysis in which phylogenetic relatedness is factored out. `cor_phylo()` does this. It can also include within-species variation (e.g., measurement error) which should better-expose the underlying correlations in traits among species. Whereas `pglmm()` can be used to identify the composition of communities within a region, `cor_phylo()` can be used to assess patterns of traits among species that make up the regional species pool.

@@

In this example, the correlation between `trait1` and `trait2` is computed, and the column named `sp` in `data` identifies the species. The object `phy.sp` specifies the phylogenetic covariance matrix as a 'phylo' object from the `ape` package. `cor_phylo()` estimates the phylogenetic signal for each trait by assuming that trait evolution is given by a Ornstein-Uhlenbeck process. The term `covariates = list(trait1 ~ env)` includes the independent variable `env` for `trait1`, to remove possible confounding effects; only an intercept is estimated if no covariate is provided for a trait. Within-species variation (measurement error) is specified by `meas_errors = list(trait1 ~ me1, trait2 ~ me2)`, where `me1` and `me2` are the standard errors for `trait1` and  `trait2`, respectively, of values at the tips of the phylogenetic tree. If within-species standard errors are not provided for a given trait, the trait values are assumed to be known without error. Finally, `cor_phylo()` can perform parametric bootstrapping to give confidence intervals for all parameter estimates: correlations, phylogenetic signals, covariate coefficients, and coefficient covariances.

@@

`pglmm()` is syntactically modeled after `lmer()` and `glmer()` in `lme4`, although it allows the specification of phylogenetic covariance matrices. `pglmm()` also allows "nested" models (with block-diagonal covariance matrices) which arise when phylogenetic covariances only act within single communities, rather than among communities. Such nested models make it possible to assess whether phylogenetic relatedness affects the abundance of species within the same communities, such as whether competition between closely related species excludes one of the competitors from communities where the other is present. Nested models are structurally incompatible with the architecture of `lme4`.

There are alternative programs to `pglmm()`, although they have limitations that `pglmm()` overcomes. @hadfield2013tale use the R package `MCMCglmm` [@mcmcglmm2010] to perform phylogenetic community analyses, although they also use ASReml because its penalized quasi-likelihood (PQL) approach is computationally much faster. Hierarchical Modelling of Species Communities (HMSC-R) [@tikhonov2019joint] performs community analyses using Bayesian MCMC approaches, although it does not include nested terms. It is also possible to code specific phylogenetic community models using flexible Bayesian platforms such as WinBugs, Stan, and JAGS, although this will involve considerable programming and expertise.

@@

Joint Species Distribution Models (JSDMs) are models where the response variable is distribution (abundances or occurrences) of multiple species across sites or samples, where all species are modeled jointly, usually by allowing non-zero covariance between either species-level errors, species-level coefficients in the model, or both [@warton2015so]. `pglmm()` in `phyr` is a joint species distribution model where the (residual) dependencies among species are modeled in a way that incorporates phylogenetic relatedness. JSDMs, and Species Distribution Models (SDM) in general, have typically been focused on producing accurate predictions of how species are distributed, usually in a geographic context. However, they can also be used for making inferences about the biology of species, such as which environmental factors are important in explaining the distribution of a species or set of species, and whether traits or evolutionary history can help explain these distributions. It is this kind of inference that is the focus of the JSDM implemented in `pglmm()`. There is often a trade-off between improving predictions and making solid inferences, because increasing the complexity or flexibility of a model can improve its predictive power, but this same complexity makes it more difficult to understand what biology is being represented by the model outputs. By incorporating phylogenetic information, `pglmm()` has two uses. First, by identifying correlations that might be expected among species due to phylogeny, `pglmm()` gives better statistical properties for tests of factors underlying community composition. For example, @li2017statistical show that failure to account for phylogenetic correlations can inflate type I errors in tests for associating environmental variables and traits that underlie community composition. Second, `pglmm()` allows explicit focus on the importance of evolutionary history in structuring species assemblages, since phylogenetic covariances are explicitly estimated. This is in contrast to many other JSDMs [e.g. those in described in @wilkinson2019comparison], which attempt to estimate all pairwise species covariances without accounting for phylogeny.

Of course, the goals of solid inference and prediction are not mutually exclusive. Good prediction requires avoiding overfitting, which can be facilitated by reducing the number of parameters in the model. In some systems, it may be possible to make better predictions using a simple phylogenetic model if phylogeny is a strong predictor of species covariance, or if many species are poorly sampled and thus estimating covariances between them individually results in higher prediction variance. Using phylogeny can help closely related species share statistical strength through phylogenetic partial pooling. Ultimately, it can be powerful to use the same kind of statistical framework to do both predictive and inferential work in ecology, because it allows ecologists to smoothly move between these two goals, both of which are important in modern ecology, and more easily and quickly draw mutual insights between them.

@@

We simulated datasets to demonstrate how to use `pglmm()` and `cor_phylo()`. Details about simulations can be found in the Appendix. Our goal in this section is to provide some general ideas about the inputs and outputs of these two functions instead of testing their statistical performances or interpreting the ecological meanings of model results. For those purposes, please see the package vignettes and @ives2018book.

@@

We fitted a PGLMM that examined how a hypothetical functional trait, environmental gradient, and their interaction affect distributions of `r nspp` species across `r nsite` sites. We focused on abundance and used the default family of data distribution (Gaussian), but other distributions can also be specified by resetting the `family` argument. Phylogenetic relationships among species and site spatial autocorrelations are specified by `cov_ranef = list(species = phy, site = Vspace)` where `species` and `site` are group variables of random terms, `phy` can be a phylogeny with class `phylo` or a phylogenetic covariance matrix, `Vspace` is a variance-covriance matrix among sites. This model can also be fitted with Bayesian framework by setting `bayes = TRUE`.

@@

The summary of model results includes the model fitting method (maximum likelihood or bayesian), the model formula, log likelihood and other related statistics (AIC, BIC, DIC), estimates of variances of random terms, coefficients of fixed terms and their uncertainties.

@@

Here, we simulated two hypothetical functional traits (`trait_1` and `trait_2`) for 50 species. We set the true correlation between these two traits to be 0.7 and their phylogenetic signals to be 0.3 and 0.95, respectively.

@@

The output of `cor_phylo()` includes log-likelihood values, AIC, BIC, estimated correlation matrix of traits, estimated phylogenetic signals of traits, estimated coefficients and their uncertainties (standard errors, Z scores, and p values). In this example, the model gave good estimates of the parameters used to simulate the data. If bootstrapping was enabled by setting the `boot` argument, the lower and upper boundaries of correlations, phylogenetic signal values, and coefficients will be appended.

@@

In recent years, there has been increasing effort to apply model-based approaches in community ecology. An example is the rise of joint species distribution models. Despite the well-known importance of phylogenetic relationships in structuring species distributions and community composition, relatively few studies have incorporated phylogenetic relationships in model-based analyses of species distributions and community ecology. A potential reason is the lack of easy-to-use tools to facilitate the use of phylogenetic species-distribution modeling in ecological communities. The package `phyr` fills this gap by providing implementations of phylogenetic species-distribution models with flexible model formula syntax (`pglmm()`). It also includes other model-based functions that are useful for ecological studies such as estimating correlations among functional traits while accounting for their evolutionary history (`cor_phylo()`) and calculating community phylogenetic diversity (e.g. `psv()`) (Table \@ref(tab:mainFunc)).

The model formula of `pglmm()` is general and can be applied using other tools to fit phylogenetic species-distribution models. Thus, `pglmm()` can serve the developer community as a shell for new methods that fit GLMMs, with `phyr` providing an easy user interface. Using INLA as a backend to fit a Bayesian version of the pglmm model is an example of this approach. To facilitate this end, we are developing phyr openly on github and actively encourage community contribution.We hope that the `phyr` package will help current and future researchers formulate and analyze phylogenetic species-distribution models.

@@

No data were used in this study. `phyr` is available at Github (https://github.com/daijiang/phyr) and CRAN (https://cran.r-project.org/web/packages/phyr/index.html).
